\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity in Human Brains}

### 2014-01-17 docs/paper/paper final.tex
veel kleine verbeteringetjes aan de hand van feedback van een proeflezer (credit to Petra)

% Title
\title{\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity in Human Brains}}
\author{
  \textbf{R. Janssen} - \href{mailto:ramon.janssen@student.ru.nl}{ramon.janssen@student.ru.nl} \\
  \textbf{T. de Ruijter} - \href{mailto:t.deruijter@student.ru.nl}{t.deruijter@student.ru.nl}\\
%  \textbf{T. Claassen} - \href{mailto:tomc@cs.ru.nl}{tomc@cs.ru.nl}\\
%  \textbf{M. Hinne} - \href{mailto:mhinne@cs.ru.nl}{mhinne@cs.ru.nl}
}

@@

\begin{abstract}
Many different approaches have been used to infer brain connectivity.
But few general causal discovery methods have yet been used to find effective connectivity.
In this article, we describe how the PC algorithm is used on functional brain-data.
This approach results in connectivity matrices which clearly show some well-known brain connections.
Some disadvantages of the PC algorithm for this use are discussed, and an attempt is made to decrease these disadvantages by making adjustments to the algorithm.
Causal relations have only be found with very low certainty. 
%In this work we applied PC algorithm on resting-state data of healthy human subjects.
%Additionally we propose a variant of PC algorithm that poses weaker assumptions on the model, better fitting the underlying model.
%We find that detecting structure homological areas in the brain works better with PC algorithm than most conventional diffusion weighted imaging techniques.
%In the area of directionality and causality, more research is needed.
\end{abstract}

\section{Introduction}
% Subject introduction
The brain and particularly the human brain have been studied for hundreds of years.
Today, brain research is still one of the thriving fields of research.
A lot has changed since the time of Phrenology, the pseudoscience attempting to derive cognitive ability and personality from measurements of the skull or, post-mortem, the brain.
Advanced and non-intrusive measurement methods exist that allow scientists to peek inside brains that give a coarse but broad overview, with widespread established clinical and research purposes.

@@

\subsection{Problem statement}
%See whether applying causal methods (PC and EMS-PC) can help bring new insights and whether we can find evidence for the existence of causal patterns on a whole brain level.
This work seeks to utilise the strengths of an established causal discovery method, namely the PC algorithm, by applying it to resting-state brain fMRI data of the human brain.
The PC algorithm can cope with hundreds of variables while remaining tractable.
We evade the problem of HRF by assuming our data comes from a distribution in a state of equilibrium, where temporal components are less strong.

We introduce an adaptation to the PC algorithm named EMS-PC that is also theoretically sound and performs at least as good as PC without additional run-time complexity.
We wonder whether application of both methods provides a means of finding consistent, specific and anatomically plausible connectivity.
As there is no baseline to compare results with, we compare these methods with another method for finding structural connectivity applied on the same resting-state data.

@@

\subsection{Execution time of the algorithm}
To apply the algorithms on our data, we have used a Xeon E5-2670, running on 2,6GHz.
To run one iteration of the PC algorithm for one subject, modified PC needed approximately one minute and EMS-PC needed approximately two minutes to find both structure and direction.
Although PC is not designed to be multithreaded, running multiple iterations lends itself for using multiple threads, as the iterations are fully independent, and only need to be averaged out afterwards.

>>>>

% Title
\title{\fontfamily{phv}\selectfont{Causal Discovery methods for Effective Connectivity in Human Brains}}
\author{
  \textbf{R. Janssen} - \href{mailto:ramon.janssen@student.ru.nl}{ramon.janssen@student.ru.nl} \\
  \textbf{T. de Ruijter} - \href{mailto:t.deruijter@student.ru.nl}{t.deruijter@student.ru.nl}\\
  \small Supervised by\\
  \small T. Claassen - \href{mailto:tomc@cs.ru.nl}{tomc@cs.ru.nl}\\
  \small M. Hinne - \href{mailto:mhinne@cs.ru.nl}{mhinne@cs.ru.nl}
}

@@

\begin{abstract}
Many different approaches have been used to infer brain connectivity.
But few general causal discovery methods have yet been used to find effective connectivity.
In this article, we describe how the PC (Peter-Clark) algorithm is used on functional brain-data.
This approach results in connectivity matrices which clearly show some well-known brain connections.
Some disadvantages of the PC algorithm for this use are discussed, and an attempt is made to decrease these disadvantages by making adjustments to the algorithm.
Causal relations have only been found with very low certainty. 
%In this work we applied PC algorithm on resting-state data of healthy human subjects.
%Additionally we propose a variant of PC algorithm that poses weaker assumptions on the model, better fitting the underlying model.
%We find that detecting structure homological areas in the brain works better with PC algorithm than most conventional diffusion weighted imaging techniques.
%In the area of directionality and causality, more research is needed.
\end{abstract}

\section{Introduction}
% Subject introduction
The brain and particularly the human brain have been studied for hundreds of years.
Today, brain research is still one of the thriving fields of research.
A lot has changed since the time of Phrenology, the pseudoscience attempting to derive cognitive ability and personality from measurements of the skull or, post-mortem, the brain.
Advanced and non-intrusive measurement methods exist that allow scientists to peek inside brains, giving a coarse but broad overview, with widespread established clinical and research purposes.

@@

\subsection{Problem statement}
%See whether applying causal methods (PC and EMS-PC) can help bring new insights and whether we can find evidence for the existence of causal patterns on a whole brain level.
This work seeks to utilise the strengths of an established causal discovery method, by applying it to resting-state brain fMRI data of the human brain.
The algorithm to be used is the the PC algorithm, named after \textbf{P}eter Spirtes and \textbf{C}larck Glymour, who developed it.
The PC algorithm can cope with hundreds of variables while remaining tractable.
We evade the problem of HRF by assuming our data comes from a distribution in a state of equilibrium, where temporal components are less strong.

We introduce an adaptation of the PC algorithm named EMS-PC that is also theoretically sound and performs at least as good as PC without additional run-time complexity.
We wonder whether application of both methods provides a means of finding consistent, specific and anatomically plausible connectivity.
As there is no baseline to compare results with, we compare these methods with DWI-measurements found by Hinne et al.\cite{hinne2013} for finding structural connectivity applied on the same resting-state data.

@@

\subsection{Execution time of the algorithm}
To apply the algorithms on our data, we have used a Xeon E5-2670, running on 2,6\emph{GHz}.
To run one iteration of the PC algorithm for one subject, modified PC needed approximately one minute and EMS-PC needed approximately two minutes to find both structure and direction.
Although PC is not designed to be multithreaded, running multiple iterations lends itself for using multiple threads, as the iterations are fully independent, and only need to be averaged out afterwards.
